---
title: ""
author: "Dominik Bahlburg"
date: "January 8, 2017"
output: 
  html_document:
    toc: true
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### A short introduction to the package ggplot2

What is ggplot?: 

* package written by Hadley Wickham 
* based on 'Grammar of Graphics' by [Wilkinson 2005](https://www.amazon.de/Grammar-Graphics-Statistics-Computing/dp/0387245448) - a tool that enables us to concisely describe the components of a graphic
* idea: layer-based system that starts with the raw data and adds stats, annnotations etc. as new layers to the plot
* or in Wickham's words: [“In brief, the grammar tells us that a statistical graphic is a mapping from data to aesthetic attributes (colour, shape, size) of geometric objects (points, lines, bars). The plot may also contain statistical transformations of the data and is drawn on a specific coordinates system.”](http://moderngraphics11.pbworks.com/f/ggplot2-Book09hWickham.pdf)
* useful links for ggplot:
** [beautiful plotting with ggplot](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/#use-a-new-theme-theme_xx). This tutorial includes a list of all properties that can be defined the the "theme"-function. It will help you to edit or even create your own theme in ggplot!
** [adding a second y-axis](http://docs.ggplot2.org/dev/sec_axis.html). This is part of the ggplot2-documentation and shows how to add a second y-axis to your plots
** [official ggplot cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf). The ggplot cheatsheet which is also uploaded to the repository and available in RStudio under Help -> Cheatsheets...
** [Tidy Data](http://vita.had.co.nz/papers/tidy-data.pdf). This is the paper by Wickham, which tells you more about the concept of "tidy data", which facilitates plotting with ggplot!

#### How to make a basic plot:  
```{r 1, echo=T, fig.width=5, fig.height=3}
library(ggplot2)
iris_test <- iris
head(iris_test)
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colours = factor(Species)))
```
  
As you can see, you don't see anything... The reason is that so far we only specified **WHAT** we want to plot (aesthetics) and not **HOW** we want to plot it (geometric objects such as dots, lines, polygons or whatever). We do this by adding a **geom_** - layer:

##### Lineplot
```{r 2, echo=T, fig.width=5, fig.height=3}
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_line()
```
  
##### Boxplot 
```{r 3, echo=T, fig.width=5, fig.height=3}
ggplot(data=iris_test,aes(x=Species, y=Sepal.Width, fill = factor(Petal.Width))) +
  geom_boxplot()
```
  
##### Scatterplot
```{r 4, echo=T, fig.width=5, fig.height=3}
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species), size = Petal.Width)) +
  geom_point() +
 labs(x="Sepal Length (cm)", y="Sepal Width (cm)", col="Species", title="Iris Dataset")
```
  
##### Heatmap
```{r 5, echo=T, fig.width=5, fig.height=3}
ggplot(data=faithfuld,aes(x=waiting, y=eruptions)) +
  geom_raster(aes(fill=density),interpolate=T) 
```  
  
##### Facet plots  
```{r 12, echo=T, fig.width=5, fig.height=3}
#facet_wrap allows us to plot our data in seperate subplots, sorted by a factor. 
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point() +
  facet_wrap(~Species)
```  
  
#### How to change  
#####colours
Not happy with the colours?
There are multiple ways to change that: 
```{r 6, echo=T, fig.width=5, fig.height=3}
#Define a vector of colours (in this case hex-codes to define them very precisely)
myColors <- c("#56ddc5", "#ff3db7", "#4699dd")
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_point() +
  scale_color_manual(values=myColors)

#Use scale_colour_hue to create a colour gradient
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_point() +
  scale_colour_hue(h=c(0, 90))

#You can also define a gradient by hand...
#What is an advantage with this method is that you can define the colours more precisely via hex-codes
colfunc <- colorRampPalette(c("#ff6f1d", "#7077df"))
myColors <- colfunc(3)
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_point() +
 scale_color_manual(values=myColors)

#or simply manipulate the lightness (l) and chroma/intensity (c) of the used colours
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_point() +
 scale_color_hue(l=50, c=60)
```
  
##### size 
  
```{r 7, echo=T, fig.width=5, fig.height=3}
#Change the symbol size to a fixed value
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species),size=3)) +
  geom_point() 

#Change the symbol size according to a value
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species),size=Petal.Length)) +
  geom_point() 
```

##### shapes
```{r 8, echo=T, fig.width=5, fig.height=3}
#Change the symbol shape to a fixed one
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_point(shape=12)

#Change the symbol shape according to a factor
ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point(shape=factor(iris_test$Species)) 
```
  
And there are countless other ways to achieve that...
Check the ggthemr and ggthemes packages which facilitate making major changes in your plot layout.
  
  
  
#### Include stats in your plots:  
##### Add linear regression to plot
```{r 9, echo=T, fig.width=5, fig.height=3}
#In case we want to add linear regression models to our scatterplot, simply add geom_smooth() to the ggplot. What we see as a grey shading around the regression lines is the predicted standard error. You can avoid that by adding se=F as an argument to geom_smooth
ggplot(iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_point() + 
  stat_smooth(method='lm')
```
  
##### Adding mean lines  
```{r 10, echo=T, fig.width=5, fig.height=3, message=F}
#
library(dplyr)
means_iris <- iris_test %>%
  group_by(.,Species)%>%
  summarise(.,mean_sw=mean(Sepal.Width),sd_sw=sd(Sepal.Width))

ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point() +
  facet_wrap(~Species) +
  geom_hline(aes(yintercept=mean_sw),data=means_iris,colour='red')
```  
  
##### mean+errorbar-plots  
```{r 13, echo=T, fig.width=5, fig.height=3, message=F}
#
ggplot(data=means_iris,aes(x=Species, y=mean_sw,colour=factor(Species))) +
  geom_point() +
  geom_errorbar(aes(ymin = mean_sw-sd_sw, ymax=mean_sw+sd_sw),width=.3) 
```  
  
#### How to add data from 2 different dataframes
```{r 14, echo=T, fig.width=5, fig.height=3, message=F}
#you can add the second dataset as a geom_point-layer
#problem: works only if both datasets have the same lengths
data1 <- as.data.frame(cbind(c(-20:20),c(-20:20)^2))
data2 <- as.data.frame(cbind(c(-20:20),c(-20:20)*3+200))

ggplot(data=data1,aes(x=V1, y=V2)) +
  geom_point() +
  geom_point(aes(x = data2$V1,y=data2$V2),colour=2)

#for cleaner code try to combine the dataframes if possible
#advantage: automatically generated legend, datasets can differ in length...
data1 <- as.data.frame(cbind(c(-20:20),c(-20:20)^2))
data2 <- as.data.frame(cbind(c(-18:18),c(-18:18)*3+200))
data3 <- rbind(data1,data2)
data3$source <- as.factor(c(rep('data1',length(data1$V1)),rep('data2',length(data2$V1))))
ggplot(data=data3,aes(x=V1,y=V2,colour=source))+
  geom_point()
```  
  
  
#### How to save your plot
To make your work easily reproducible you should try to avoid saving your plots via clicking "Export","Save Plot"...  
A better solution is to use functions such as ggsave!
```{r 11, echo=T, fig.width=5, fig.height=3}
plot1 <- ggplot(data=iris_test,aes(x=Sepal.Length, y=Sepal.Width, colour = factor(Species))) +
  geom_point()
ggsave('plot1.png',plot=plot1,width=6,height=4)
```
This saves your plot "plot1" as a png-file (or whatever format you want) in your current working directory. You can also specify a specific directory where the plot should be saved (e.g. ggsave('/Users/maxmustermann/plot1.png',...). The arguments width and height determine the width and height of the plot (which is 600x400 pixel in our case).
  
#### Some cool stuff
```{r 16, message=F, warning=F}
sites <- read.csv("sites.csv", sep=" ", dec=".")
#This is just to show you what ggplot is capable of doing...
#Here we plot a simple map via ggmap to quickly visualize a few sites where samples for a study were taken
#Additionally, we use the package ggrepel to avoid overlapping of the station labels. This is extremely helpful since ggplot is not able to do that by itself. 
library(ggrepel)
library(ggmap)
#Define borders for the map (min lon, min lat, max lon, max lat)
antarctica_coords<-c(-62, -65, -52, -60)
#we download a map from the google server. You can choose between multiple sources such as OpenStreetMap...
antarctica<-get_map(location=antarctica_coords,source="google",maptype="roadmap", crop=FALSE)

ggmap(antarctica)+
  geom_point(data=sites,aes(x=Longitude, y=Latitude),cex=0.25)+
  geom_text_repel(data=sites,aes(x=Longitude,y=Latitude,label = station,vjust=0.1,force=2)) +
  labs(x='Longitude',y='Latitude')
```